---
title: "ST558 - Project 1"
author: "Yefrid Cordoba"
format: html
---

## Introduction



## 1. Data Processing

### 1.1 Parsing Variable Values

First we start by loading the libraries that we need

```{r}
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
library(hms)
library(janitor)
```


Preparing the data we start downloading and parsing the JSON files with the variables description.

To perform this job we need to create a link that it is needed to request the variables info from the census API, as year is one of our variables, we use the function `paste0()` to add the year that is going to be used by the function.   
Then we feed the link to the function `GET()` and parse the data from the JSON file.   
We get the variables that we need (Shown below in the function definition), applying the correct modification for `"ST"` when the year is 2021 or 2022.

```{r}
var_census <- function(year){
  link <- paste0("https://api.census.gov/data/", as.character(year),
                 "/acs/acs1/pums/variables.json")
  
  census_var <- GET(url = link)
  
  var_parsed <- fromJSON(rawToChar(census_var$content))[[1]]
  
  state <- "STATE"
  
  if(year %in% c(2021, 2022)){
    state <- "ST"
  }
  
  var_to_keep <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "HHL", "SCH", "SCHL", "SEX", state, "REGION",
                   "DIVISION")
  var_parsed <- var_parsed[var_to_keep]
  
  return(var_parsed)

}
```

Now we need to get the variables for each year, and save them in a list that we will save in a `.rds` file.

```{r}
Variables_21_24 <- list(
  Variables_21 = var_census(2021),
  Variables_22 = var_census(2022),
  Variables_23 = var_census(2023),
  Variables_24 = var_census(2024)
)

saveRDS(Variables_21_24, file = "data/Variables_census.rds")
```

### 1.2 Parsing API Data

Start by getting the usual process to work with a given URL??

We are going to write a helper function that takes what is returned by `GET()` and turn it into a nice tibble.

```{r}
get_to_tibble <- function(u_info){
  
  parsed <- fromJSON(rawToChar(u_info$content))
  
  n_tibble <- as_tibble(parsed) |>
    row_to_names(row_number = 1)
  
  return(n_tibble)
  
}
```

Now we are going to create a function to query the census API, where the user specifies year, the numeric and categorical variables to be returned, the geography level, and the specific region, division or state to subset the data on.

We are going to use this helper function to handle the time variables (created using LLM).

```{r}

parse_jwp_midpoint <- function(jwp_vector) {
  # 1. Regex pattern to extract ONLY valid time range substrings
  # Ignores prefixes like "001: " and returns NA for non-time entries like "0: N/A..."
  pattern <- "\\d{1,2}:\\d{2}\\s*(?:a\\.m\\.|p\\.m\\.|am|pm)\\s+to\\s+\\d{1,2}:\\d{2}\\s*(?:a\\.m\\.|p\\.m\\.|am|pm)"
  time_ranges <- stringr::str_extract(jwp_vector, pattern)
  
  # 2. Clean dots from a.m./p.m.
  clean_vec <- gsub("\\.", "", time_ranges)
  
  # 3. Split into start and end time columns
  times_matrix <- stringr::str_split_fixed(clean_vec, " to ", 2)
  
  # 4. Parse times into POSIXct objects
  t_start <- parse_date_time(times_matrix[, 1], orders = "I:M p", quiet = TRUE)
  t_end   <- parse_date_time(times_matrix[, 2], orders = "I:M p", quiet = TRUE)
  
  # 5. Handle overnight ranges (e.g., 11:50 p.m. to 12:10 a.m.)
  overnight <- !is.na(t_start) & !is.na(t_end) & (t_end < t_start)
  t_end[overnight] <- t_end[overnight] + days(1)
  
  # 6. Calculate midpoint and convert to time class
  midpoint <- t_start + (t_end - t_start) / 2
  as_hms(midpoint)
}
```


```{r}
parse_census <- function(user_year = 2024, num_var = c("AGEP", "PWGTP"), 
                         cat_var = c("SEX"), geo = "STATE", sub_geo = NULL ){
  #Checking for the correct year
  user_year <- as.numeric(user_year)
  if(!(user_year %in% c(2021, 2022, 2023, 2024)) || length(user_year) != 1){
    stop("Year should be one of (2021, 2022, 2023, or 2024)")
  }
  
   geo_api <- tolower(geo)
  # 
  # if(user_year %in% c(2021, 2022)){
  #   geo <- "ST"
  # }
  # 
  #Turning every input in upper case to standardize
  num_var <- toupper(as.character(num_var))
  cat_var <- toupper(as.character(cat_var))
  geo <- toupper(as.character(geo))
  
  
  
  #Checking for the presence of "PWGTP", always needs to be returned
  if(!("PWGTP" %in% num_var)){
    num_var <- c(num_var, "PWGTP")
  }
  
  #Checking that also another numeric variable is shown
  if(length(num_var) == 1 && num_var == "PWGTP"){
    print("Needs to specify a numric variable other than `PWGTP`, `AGEP` will be added by default")
    num_var <- c("AGEP", num_var)
  }
  
  #Checking that the numeric variables are in the allowed values
  num_allowed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!all(num_var %in% num_allowed)){
    stop("Numeric variables need to be between (AGEP, GASP, GRPIP, JWAP (time), JWDP (time), and JWMNP")
  }
  
  #Checking for the categorical variables
  cat_allowed <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(!all(cat_var %in% cat_allowed))
    stop("Categorical variables need to be between (FER, HHL, SCH, SCHL, and SEX)")
  
  #Checking that the geography level is just one of the geographies allowed
  geo_allowed <- c("REGION", "DIVISION", "STATE", "ST")
  if(!(geo %in% geo_allowed) || length(geo) != 1){
    stop("Geography level needs to be ONE between REGION, DIVISION, or STATE")
  }  
    
  #Variable for the API query, which always use "state" in the query.
  geo_api <- tolower(geo)
  
  #Variable for the stored data
  geo_var <- if(user_year %in% c(2021, 2022) && geo == "STATE") {
    "ST"
    } else {
    geo
    }
    
  # Checking that the allowed variables are correct
  year_2digit <- substr(as.character(user_year), 3, 4)
  
  #Get it to look in the correct list of variables
  target_name <- paste0("Variables_", year_2digit)
  Variables <- readRDS("data/Variables_census.rds")
  
  variables_m <- Variables[[target_name]]
  
  
  #Lower case to filter the api
  #geo_api <- tolower(geo)
  
  #Assigning default values for each of the geography levels
  if(is.null(sub_geo)){
    sub_geo <- switch(
      geo,
      STATE = "North Carolina/NC",
      ST = "North Carolina/NC",
      REGION = "South",
      DIVISION = "East South Central (South region)"
    )
  }
  
  
  #print(paste0("after", geo))

  
  if(all(tolower(sub_geo) != "all") && 
     !all(sub_geo %in% variables_m[[geo_var]]$values$item)){
    
    stop(paste0("Geography subset is not valid for ",geo))
  }
  
  #Turning inputs in census codes
  geo_lookup <- variables_m[[geo_var]]$values$item
  #print(geo_lookup)
  
  #print(sub_geo)
  
  if(all(tolower(sub_geo) == "all")){
    geo_codes <- "*"
  } else {
      geo_codes <- names(geo_lookup[match(sub_geo, geo_lookup)])
  
      geo_codes <- as.character(geo_codes)
      
      geo_codes <- paste(geo_codes, collapse = ",")
  }
  #print(geo_codes)
  #Putting the columns into a string for the query
  all_col <- unique(c(num_var, cat_var))
  
  string_cols <- paste(all_col, collapse = ",")
  
  census_key <- Sys.getenv("census_key")
  
  #As the api just allow to request one geography or all geographies, we will create a helper for one geography and then mix the outputs.

  census_response <- httr::GET(
    url = paste0("https://api.census.gov/data/",as.character(user_year),
               "/acs/acs1/pums"),
    query = list(
    get = string_cols,
    `for` = paste0(geo_api, ":", geo_codes),
    key = census_key
    )
    )
  #census_response
  census_data <- get_to_tibble(census_response) 
  
  #Changing the categorical variables into factors
  for (one_cat in cat_var){
    
    #Get the list that we are going to use to look up
    cat_lookup <- variables_m[[one_cat]]$values$item
    
    #Obtain a names vector to look for the variables that match
    cat_lookup <- unlist(cat_lookup,
                         use.names = TRUE)
    
    #Lookup and replace the values
    census_data[[one_cat]] <- factor(
      unname(cat_lookup[as.character(census_data[[one_cat]])])
    )
  }
    
  #Changing the time variables into theit midpoint
  
  time_var <- intersect( c("JWAP", "JWDP"),
                         names(census_data))
  
  #We are going to iterate to get the actual string based on the code provided
  for (one_time in time_var){

    #Getting the list to lookup
    time_lookup <- variables_m[[one_time]]$values$item

    #Turning them int a named vector
    time_lookup <- unlist( time_lookup,
                           use.names = TRUE)

    #Putting the time string before using the midpoint function helper
    time_string <- unname( time_lookup[as.character(census_data[[one_time]])]
                           )

    #Now we can rename the column with the codes and use the midpoint function
    census_data[[one_time]] <- parse_jwp_midpoint(time_string)
  }
  
  #Getting the values that are different from the time variables
  numeric_var <- setdiff( num_var,
                          c("JWAP", "JWDP"))
  
  #Changing them to numric variables
  census_data <- census_data |>
    mutate(
      across(
        any_of(numeric_var),
        as.numeric
      )
    )
  
  #And we want to get the labels for the geography data
  geo_column <- geo_api
  
  #We also need to get the correct value to look for in the variables (different for 2021 and 2022)
  geo_lookup <- variables_m[[geo_var]]$values$item
  
  #We get the named vector
  geo_lookup <- unlist(geo_lookup, use.names = TRUE)
  
  #and we put the names in the tibble
  census_data[[geo_column]] <- factor(
    unname( geo_lookup[as.character(census_data[[geo_column]])])
  )
  
  #Required code
  class(census_data) <- c("census", class(census_data))
  
  census_data

}
```

### 1.3 Combining Years

Now we are going to create a function that will use the previous function for multiple years and will append those tibbles in a new tibble with the specified year.

```{r}
parse_census_years <- function(user_years = c(2022, 2024),
                               num_var = c("AGEP", "PWGTP"),
                               cat_var = c("SEX"),
                               geo = "STATE",
                               sub_geo = NULL){
  
  #Checking that the years are allowed
  user_years <- as.numeric(user_years)
  
  if( !all(user_years %in% c(2021, 2022, 2023, 2024))){
    stop ("Years should be between 2021, 2022, 2023, or 2024")
  }
  
  #Create a list for the results
  census_list <- list()
  
  #Now we use a loop for each year
  for (one_year in user_years){
    
    #applying the funstion prevously created for each year
    census_year <- parse_census(
      user_year = one_year,
      num_var = num_var,
      cat_var = cat_var,
      geo = geo,
      sub_geo = sub_geo
    )
    
    #Now we add the year variable
    census_year <- census_year |>
      mutate(year = one_year)
    
    #and finally we save the tibble in the list
    census_list[[as.character(one_year)]] <- census_year
    
  }
  
  #Finally we combine all the results from the list of tibbles into a tibble
  census_all <- bind_rows(
    census_list) |>
    relocate(year)
  
  return(census_all)
}
```

### 1.4 Testing

First we are going to check that the restrictions work.   

The first one is check that is for the correct years.

```{r}
#| error: TRUE

parse_census_years(2020)

#For a different numeric variable than the allowed
parse_census_years(2021, num_var = "AGS")

#Trying to get more than one geography
parse_census_years(2022, geo = c("REGION", "STATE"))

#Requesting geographies that are not part of the one chosen
parse_census_years(2022, geo = "STATE", sub_geo = "South")
```

Now we are going to extract data that is valid and get the tibble for those.

Using the smallest state (by population) to make sure it runs smoothly.

```{r}
WY_data <- parse_census_years(2021:2023, num_var = c("AGEP", "GASP", "JWAP"), 
                   cat_var = c("FER", "SEX"), geo = "STATE", 
                   sub_geo = "Wyoming/WY")
WY_data
```


## 2. Writing a Generic Function for Summarizing

### 2.1 Numeric Summaries

For census summary method we are going to write a function that produces means and standard deviations for our numeric variable, and counts for the categorical variables.

```{r}
summary.census <- function(census_tibble, num_var = NULL, cat_var = NULL){
  
  #Getting all the summaries for the numeric variables when num_var is not especified
  if( is.null(num_var)){
    
    #First we select all the variables that are numeric
    num_var <- names(census_tibble)[sapply(census_tibble, is.numeric)]
    
    #And remove the variables that we cannot do summaries with
    
    num_var <- setdiff(num_var, c("PWGTP", "year", "JWAP", "JWDP"))
    
  } else{
    
    #we standardize the input
    num_var <- toupper(as.character(num_var))
    
    #And remove the time variable which cannot be summarized
    num_var <- setdiff(num_var, c("PWGTP", "JWAP", "JWDP"))
    
  }
  
  #And we are going to do the same for the categorical variables
  if(is.null(cat_var)){
    
    #Get the categorical variables
    cat_var <- names(census_tibble)[sapply(census_tibble,
                                          function(x){is.factor(x) ||
                                              is.character(x)})]
    
    #Year needs to be treated as categorical
    if("year" %in% names(census_tibble) && !("year" %in% cat_var)){
      
      cat_var <- c(cat_var, "year")
    }
  } else {
    
    cat_var <- as.character(cat_var)
    
   #Now we are going to correct the case for the geography and year variables from the tibble 
    for (i in seq_along(cat_var)){
      
      if(!(cat_var[i] %in% names(census_tibble)) &&
         toupper(cat_var[i]) %in% names(census_tibble)){
        
        cat_var[i] <- toupper(cat_var[i])
      }
    }
  }
  
  #Checking that the requested variable exist
 
  num_not_found <- setdiff(num_var,
    names(census_tibble))

  if (length(num_not_found) > 0) {
    stop("The numeric variables were not found")
  }

  cat_not_found <- setdiff(cat_var,
    names(census_tibble))

  if (length(cat_not_found) > 0) {
    stop("The categorical variables were not found")
  }
  
  #For the output, we are going to store our results in lists
  num_summaries <- list()
  cat_summaries <- list()

  #We are going to calculate the numeric summaries
  for (one_num_var in num_var) {

    numeric_vector <- census_tibble[[one_num_var]]
    weight_vector <- census_tibble[["PWGTP"]]

    # Keeping only observations where the variable and weight exist
    complete_observations <- (!is.na(numeric_vector) &
      !is.na(weight_vector))

    numeric_vector_complete <- numeric_vector[complete_observations]

    weight_vector_complete <- weight_vector[complete_observations]

    #We need to check that observations exist where both of the values match
    if (
      length(numeric_vector_complete) == 0 ||
      sum(weight_vector_complete) == 0
    ) {

      sample_mean <- NA_real_
      sample_sd <- NA_real_

    } else {

      # After confirming that, we proceed to calculate the summaries
      sample_mean <- sum(numeric_vector_complete *
                           weight_vector_complete) / 
        sum(weight_vector_complete)

      # Weighted sample standard deviation
      sample_sd <- sqrt( sum(numeric_vector_complete^2 *
                               weight_vector_complete) /
                           sum(weight_vector_complete) -
                           sample_mean^2)
    }

    # Save the results for the variable in the list
    num_summaries[[one_num_var]] <- c(mean = sample_mean,
                                      standard_deviation = sample_sd)
  }
  
  #And we calculate the values for the categorical variables
  for (one_cat_var in cat_var) {

    categorical_vector <- census_tibble[[one_cat_var]]

    # Year must be treated as categorical
    if (one_cat_var == "year") {
      categorical_vector <- factor(categorical_vector)
    }

    # Count the observations in every category
    category_counts <- table(categorical_vector, useNA = "ifany")

    cat_summaries[[one_cat_var]] <- category_counts
  }
  
  #At the end we put everything in a list with all the summaries
  census_summary <- list(
    numeric_summaries = num_summaries,
    categorical_counts = cat_summaries
  )

  return(census_summary)
  
}
```

### 2.2 Graphical Summaries

In this section we are going to create a generic plot() function for a census class tibble.
It is required that the user specify one categorical and one numeric variable. The output will be a boxplot of the numeric variable at the levels of the categorical variable.

```{r}
plot.census <- function(census_tibble, cat_var, num_var,...) {

  # Turning the variable names into character values
  cat_var <- as.character(cat_var)
  num_var <- as.character(num_var)

  # Checking that only one categorical variable was provided
  if (length(cat_var) != 1) {
    stop("You need to specify exactly one categorical variable.")
  }

  # Checking that only one numeric variable was provided
  if (length(num_var) != 1) {
    stop("You need to specify exactly one numeric variable.")
  }

  # If the user writes the variable names in lowercase, check whether the uppercase version exists
  if(!(cat_var %in% names(census_tibble)) &&
    toupper(cat_var) %in% names(census_tibble)) {
    cat_var <- toupper(cat_var)
  }

  if(!(num_var %in% names(census_tibble)) &&
    toupper(num_var) %in% names(census_tibble)) {
    num_var <- toupper(num_var)
  }

  # Checking that the categorical variable exists
  if (!(cat_var %in% names(census_tibble))) {
    stop("The categorical variable was not found in the census tibble")
  }

  # Checking that the numeric variable exists
  if (!(num_var %in% names(census_tibble))) {
    stop("The numeric variable was not found in the census tibble")
  }

  # Time variables should not be used as numeric variables
  if (num_var %in% c("JWAP", "JWDP")) {
    stop("JWAP and JWDP are time variables and cannot be used as numeric variables")
  }

  # PWGTP should only be used as the weight
  if (num_var == "PWGTP") {
    stop("PWGTP is the weight variable and cannot be used as the numeric variable")
  }

  # Checking that the selected numeric variable is numeric
  if (!is.numeric(census_tibble[[num_var]])) {
    stop(num_var, "needs to be a numeric variable")
  }

  # Treating year as a categorical variable
  if (cat_var == "year") {
    census_tibble$year <- factor(census_tibble$year)
  }

  # Also make the selected categorical variable a factor in case they are not already
  if (!is.factor(census_tibble[[cat_var]])) {
    census_tibble[[cat_var]] <- factor(census_tibble[[cat_var]])
  }
print(paste(cat_var, num_var, sep = ", "))
  # Creating the weighted boxplot using ggplot
  census_plot <- ggplot(data = census_tibble,
                        aes(x = .data[[cat_var]],
                            y = .data[[num_var]],
                            weight = PWGTP)
  ) +
    geom_boxplot(na.rm = TRUE) +
    labs(title = paste(
          num_var,
          "by",
          cat_var),
      x = cat_var,
      y = num_var
    ) +
    theme_bw()
  
  return(census_plot)
}
```

### 2.3 Testing

First we are going to test the numerical summaries.

```{r}
summary.census(WY_data)
```

And then test Graphical summaries

```{r}
plot.census(WY_data, num_var = "GASP", cat_var = "year")
```

## Investigation

