---
title: "ST558 - Project 1"
author: "Yefrid Cordoba"
format: html
---

## Introduction



## 1. Data Processing

### 1.1 Parsing Variable Values

First we start by loading the libraries that we need

```{r}
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(httr)
library(jsonlite)
```


Preparing the data we start downloading and parsing the JSON files with the variables description.

To perform this job we need to create a link that it is needed to request the variables info from the census API, as year is one of our variables, we use the function `paste0()` to add the year that is going to be used by the function.   
Then we feed the link to the function `GET()` and parse the data from the JSON file.   
We get the variables that we need (Shown below in the function definition), applying the correct modification for `"ST"` when the year is 2021 or 2022.

```{r}
var_census <- function(year){
  link <- paste0("https://api.census.gov/data/", as.character(year),
                 "/acs/acs1/pums/variables.json")
  
  census_var <- GET(url = link)
  
  var_parsed <- fromJSON(rawToChar(census_var$content))[[1]]
  
  state <- "STATE"
  
  if(year %in% c(2021, 2022)){
    state <- "ST"
  }
  
  var_to_keep <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "HHL", "SCH", "SCHL", "SEX", state, "REGION",
                   "DIVISION")
  var_parsed <- var_parsed[var_to_keep]
  
  return(var_parsed)

}
```

Now we need to get the variables for each year, and save them in a list that we will save in a `.rds` file.

```{r}
Variables_21_24 <- list(
  Variables_21 = var_census(2021),
  Variables_22 = var_census(2022),
  Variables_23 = var_census(2023),
  Variables_24 = var_census(2024)
)

saveRDS(Variables_21_24, file = "data/Variables_census.rds")
```

### 1.2 Parsing API Data

Start by getting the usual process to work with a given URL??

We are going to write a helper function that takes what is returned by `GET()` and turn it into a nice tibble.

```{r}
get_to_tibble <- function(u_info){
  
  parsed <- fromJSON(rawToChar(u_info$content))
  
  n_tibble <- as_tibble(parsed$data)
  
  return(n_tibble)
  
}
```

Now we are going to create a function to query the census API, where the user specifies year, the numeric and categorical variables to be returned, the geography level, and the specific region, division or state to subset the data on.

```{r}
parse_census <- function(year = 2024, num_var = c("AGEP", "PWGTP"), 
                         cat_var = c("SEX"), geo = "STATE", sub_geo = NULL ){
  #Checing for the correct year
  if(!(year %in% c(2021, 2022, 2023, 2024))){
    stop("Year should be one of (2021, 2022, 2023, or 2024)")
  }
  
  #Checking for the presence of "PWGTP", always needs to be returned
  if(!("PWGTP" %in% num_var)){
    num_var <- c(num_var, "PWGTP")
  }
  
  
  
}
```

