---
title: "ST558 - Project 1"
author: "Yefrid Cordoba"
format: html
---

## Introduction



## 1. Data Processing

### 1.1 Parsing Variable Values

First we start by loading the libraries that we need

```{r}
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
library(hms)
library(janitor)
```


Preparing the data we start downloading and parsing the JSON files with the variables description.

To perform this job we need to create a link that it is needed to request the variables info from the census API, as year is one of our variables, we use the function `paste0()` to add the year that is going to be used by the function.   
Then we feed the link to the function `GET()` and parse the data from the JSON file.   
We get the variables that we need (Shown below in the function definition), applying the correct modification for `"ST"` when the year is 2021 or 2022.

```{r}
var_census <- function(year){
  link <- paste0("https://api.census.gov/data/", as.character(year),
                 "/acs/acs1/pums/variables.json")
  
  census_var <- GET(url = link)
  
  var_parsed <- fromJSON(rawToChar(census_var$content))[[1]]
  
  state <- "STATE"
  
  if(year %in% c(2021, 2022)){
    state <- "ST"
  }
  
  var_to_keep <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "HHL", "SCH", "SCHL", "SEX", state, "REGION",
                   "DIVISION")
  var_parsed <- var_parsed[var_to_keep]
  
  return(var_parsed)

}
```

Now we need to get the variables for each year, and save them in a list that we will save in a `.rds` file.

```{r}
Variables_21_24 <- list(
  Variables_21 = var_census(2021),
  Variables_22 = var_census(2022),
  Variables_23 = var_census(2023),
  Variables_24 = var_census(2024)
)

saveRDS(Variables_21_24, file = "data/Variables_census.rds")
```

### 1.2 Parsing API Data

Start by getting the usual process to work with a given URL??

We are going to write a helper function that takes what is returned by `GET()` and turn it into a nice tibble.

```{r}
get_to_tibble <- function(u_info){
  
  parsed <- fromJSON(rawToChar(u_info$content))
  
  n_tibble <- as_tibble(parsed) |>
    row_to_names(row_number = 1)
  
  return(n_tibble)
  
}
```

Now we are going to create a function to query the census API, where the user specifies year, the numeric and categorical variables to be returned, the geography level, and the specific region, division or state to subset the data on.

We are going to use this helper function to handle the time variables (created using LLM).

```{r}

parse_jwp_midpoint <- function(jwp_vector) {
  # 1. Regex pattern to extract ONLY valid time range substrings
  # Ignores prefixes like "001: " and returns NA for non-time entries like "0: N/A..."
  pattern <- "\\d{1,2}:\\d{2}\\s*(?:a\\.m\\.|p\\.m\\.|am|pm)\\s+to\\s+\\d{1,2}:\\d{2}\\s*(?:a\\.m\\.|p\\.m\\.|am|pm)"
  time_ranges <- stringr::str_extract(jwp_vector, pattern)
  
  # 2. Clean dots from a.m./p.m.
  clean_vec <- gsub("\\.", "", time_ranges)
  
  # 3. Split into start and end time columns
  times_matrix <- stringr::str_split_fixed(clean_vec, " to ", 2)
  
  # 4. Parse times into POSIXct objects
  t_start <- parse_date_time(times_matrix[, 1], orders = "I:M p", quiet = TRUE)
  t_end   <- parse_date_time(times_matrix[, 2], orders = "I:M p", quiet = TRUE)
  
  # 5. Handle overnight ranges (e.g., 11:50 p.m. to 12:10 a.m.)
  overnight <- !is.na(t_start) & !is.na(t_end) & (t_end < t_start)
  t_end[overnight] <- t_end[overnight] + days(1)
  
  # 6. Calculate midpoint and convert to time class
  midpoint <- t_start + (t_end - t_start) / 2
  as_hms(midpoint)
}
```


```{r}
parse_census <- function(user_year = 2024, num_var = c("AGEP", "PWGTP"), 
                         cat_var = c("SEX"), geo = "STATE", sub_geo = NULL ){
  #Checking for the correct year
  user_year <- as.numeric(user_year)
  if(!(user_year %in% c(2021, 2022, 2023, 2024)) || length(user_year) != 1){
    stop("Year should be one of (2021, 2022, 2023, or 2024)")
  }
  
   geo_api <- tolower(geo)
  # 
  # if(user_year %in% c(2021, 2022)){
  #   geo <- "ST"
  # }
  # 
  #Turning every input in upper case to standardize
  num_var <- toupper(as.character(num_var))
  cat_var <- toupper(as.character(cat_var))
  geo <- toupper(as.character(geo))
  
  
  
  #Checking for the presence of "PWGTP", always needs to be returned
  if(!("PWGTP" %in% num_var)){
    num_var <- c(num_var, "PWGTP")
  }
  
  #Checking that also another numeric variable is shown
  if(length(num_var) == 1 && num_var == "PWGTP"){
    print("Needs to specify a numric variable other than `PWGTP`, `AGEP` will be added by default")
    num_var <- c("AGEP", num_var)
  }
  
  #Checking that the numeric variables are in the allowed values
  num_allowed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!all(num_var %in% num_allowed)){
    stop("Numeric variables need to be between (AGEP, GASP, GRPIP, JWAP (time), JWDP (time), and JWMNP")
  }
  
  #Checking for the categorical variables
  cat_allowed <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(!all(cat_var %in% cat_allowed))
    stop("Categorical variables need to be between (FER, HHL, SCH, SCHL, and SEX)")
  
  #Checking that the geography level is just one of the geographies allowed
  geo_allowed <- c("REGION", "DIVISION", "STATE", "ST")
  if(!(geo %in% geo_allowed) || length(geo) != 1){
    stop("Geography level needs to be ONE between REGION, DIVISION, or STATE")
  }  
    
  #Variable for the API query, which always use "state" in the query.
  geo_api <- tolower(geo)
  
  #Variable for the stored data
  geo_var <- if(user_year %in% c(2021, 2022) && geo == "STATE") {
    "ST"
    } else {
    geo
    }
    
  # Checking that the allowed variables are correct
  year_2digit <- substr(as.character(user_year), 3, 4)
  
  #Get it to look in the correct list of variables
  target_name <- paste0("Variables_", year_2digit)
  Variables <- readRDS("data/Variables_census.rds")
  
  variables_m <- Variables[[target_name]]
  
  
  #Lower case to filter the api
  #geo_api <- tolower(geo)
  
  #Assigning default values for each of the geography levels
  if(is.null(sub_geo)){
    sub_geo <- switch(
      geo,
      STATE = "North Carolina/NC",
      ST = "North Carolina/NC",
      REGION = "South",
      DIVISION = "East South Central (South region)"
    )
  }
  
  
  #print(paste0("after", geo))

  
  if(all(tolower(sub_geo) != "all") && 
     !all(sub_geo %in% variables_m[[geo_var]]$values$item)){
    
    stop(paste0("Geography subset is not valid for ",geo))
  }
  
  #Turning inputs in census codes
  geo_lookup <- variables_m[[geo_var]]$values$item
  #print(geo_lookup)
  
  #print(sub_geo)
  
  if(all(tolower(sub_geo) == "all")){
    geo_codes <- "*"
  } else {
      geo_codes <- names(geo_lookup[match(sub_geo, geo_lookup)])
  
      geo_codes <- as.character(geo_codes)
      
      geo_codes <- paste(geo_codes, collapse = ",")
  }
  #print(geo_codes)
  #Putting the columns into a string for the query
  all_col <- unique(c(num_var, cat_var))
  
  string_cols <- paste(all_col, collapse = ",")
  
  census_key <- Sys.getenv("census_key")
  
  #As the api just allow to request one geography or all geographies, we will create a helper for one geography and then mix the outputs.

  census_response <- httr::GET(
    url = paste0("https://api.census.gov/data/",as.character(user_year),
               "/acs/acs1/pums"),
    query = list(
    get = string_cols,
    `for` = paste0(geo_api, ":", geo_codes),
    key = census_key
    )
    )
  #census_response
  census_data <- get_to_tibble(census_response) 
  
  #Changing the categorical variables into factors
  for (one_cat in cat_var){
    
    #Get the list that we are going to use to look up
    cat_lookup <- variables_m[[one_cat]]$values$item
    
    #Obtain a names vector to look for the variables that match
    cat_lookup <- unlist(cat_lookup,
                         use.names = TRUE)
    
    #Lookup and replace the values
    census_data[[one_cat]] <- factor(
      unname(cat_lookup[as.character(census_data[[one_cat]])])
    )
  }
    
  #Changing the time variables into theit midpoint
  
  time_var <- intersect( c("JWAP", "JWDP"),
                         names(census_data))
  
  #We are going to iterate to get the actual string based on the code provided
  for (one_time in time_var){

    #Getting the list to lookup
    time_lookup <- variables_m[[one_time]]$values$item

    #Turning them int a named vector
    time_lookup <- unlist( time_lookup,
                           use.names = TRUE)

    #Putting the time string before using the midpoint function helper
    time_string <- unname( time_lookup[as.character(census_data[[one_time]])]
                           )

    #Now we can rename the column with the codes and use the midpoint function
    census_data[[one_time]] <- parse_jwp_midpoint(time_string)
  }
  
  #Getting the values that are different from the time variables
  numeric_var <- setdiff( num_var,
                          c("JWAP", "JWDP"))
  
  #Changing them to numric variables
  census_data <- census_data |>
    mutate(
      across(
        any_of(numeric_var),
        as.numeric
      )
    )
  
  #And we want to get the labels for the geography data
  geo_column <- geo_api
  
  #We also need to get the correct value to look for in the variables (different for 2021 and 2022)
  geo_lookup <- variables_m[[geo_var]]$values$item
  
  #We get the named vector
  geo_lookup <- unlist(geo_lookup, use.names = TRUE)
  
  #and we put the names in the tibble
  census_data[[geo_column]] <- factor(
    unname( geo_lookup[as.character(census_data[[geo_column]])])
  )
  
  #Required code
  class(census_data) <- c("census", class(census_data))
  
  census_data

}
```

